Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-23 ind_files/01_Introduction.tex
Format citations in epigraphs

\epigraph{\justifying The theory of evolution by natural selection is an ecological theory—founded on ecological observation by perhaps the greatest of all ecologists. It has been adopted by and brought up by the science of genetics, and ecologists, being modest people, are apt to forget their distinguished parenthood}{John Harper~\citep{harper_darwinian_1967}}

@@

The most general mathematical framework we have for evolutionary population biology is the Price equation~\citep{frank_natural_2012,frank_universal_2017,queller_fundamental_2017,luque_one_2017,lion_theoretical_2018,lehtonen_price_2018,lehtonen_price_2020,luque_mirror_2021}. Indeed, much like statistical mechanics in physics, the Price equation is derived from a small number of very general first principles and is able to recover several standard equations as special cases. The Price equation partitions changes in population composition into multiple terms, each of which lends itself to a straightforward interpretation in terms of `high-level' evolutionary forces such as selection and mutation, thus providing a useful conceptual framework for thinking about how populations change over time~\citep{frank_natural_2012}. However, the greater complexity of ecology and evolution relative to physics has meant that the generality of the Price equation comes with a much bigger cost in predictive power. the Price equation in its most general formulation is dynamically insufficient\footnote{Since this is not very standard nomenclature outside theoretical biology and related fields: `Dynamically insufficient' means that the equation cannot be iterated, or presented in the form of a difference equation $x_{t+1} = F(x_t)$ or a differential equation $\dot{x} = F(x)$. Thus, we cannot predict a dynamic `trajectory' given an initial condition $x_0$. The term `insufficient' is to indicate that the equation requires `complete' information to be true and only relates two quantities `in retrospect'. It is thus `insufficient' for prediction. In our case, in its most general setting, the Price equation is usually formulated in a way that partitions a \emph{given} amount of phenotypic change between two populations (usually but not necessarily the same population at two different times) into change due to selection, transmission bias, etc., rather than \emph{predicting} a trajectory for how much phenotypic change will occur at various future times based on the \emph{current} phenotypic distribution.}~\citep{van_veelen_use_2005,frank_natural_2012,simon_continuous-time_2014,queller_fundamental_2017}. However, this need not be the case if we are willing to compromise slightly - Several authors have put forth more predictive versions of the Price equation by moving to a continuous time differential equation framework in which the Price equation is dynamically sufficient but manifests in a slightly less general form~\citep{page_unifying_2002,lion_theoretical_2018,day_price_2020}.

>>>>

\epigraph{\justifying The theory of evolution by natural selection is an ecological theory—founded on ecological observation by perhaps the greatest of all ecologists. It has been adopted by and brought up by the science of genetics, and ecologists, being modest people, are apt to forget their distinguished parenthood}{John~\citet{harper_darwinian_1967}}

@@

The most general mathematical framework we have for evolutionary population biology is the Price equation~\citep{frank_natural_2012,frank_universal_2017,queller_fundamental_2017,luque_one_2017,lion_theoretical_2018,lehtonen_price_2018,lehtonen_price_2020,luque_mirror_2021}. Indeed, much like statistical mechanics in physics, the Price equation is derived from a small number of very general first principles and is able to recover several standard equations as special cases. The Price equation partitions changes in population composition into multiple terms, each of which lends itself to a straightforward interpretation in terms of `high-level' evolutionary forces such as selection and mutation, thus providing a useful conceptual framework for thinking about how populations change over time~\citep{frank_natural_2012}. However, the greater complexity of ecology and evolution relative to physics has meant that the generality of the Price equation comes with a much bigger cost in predictive power. the Price equation in its most general formulation is dynamically insufficient\footnote{Since this is not very standard nomenclature outside theoretical biology and related fields: `Dynamically insufficient' means that the equation cannot be presented in the form of a difference equation $x_{t+1} = F(x_t)$ or a differential equation $\dot{x} = F(x)$. Thus, we cannot predict a dynamic `trajectory' given an initial condition $x_0$. The term `insufficient' is to indicate that the equation requires `complete' information to be true and only relates two quantities `in retrospect'. It is thus `insufficient' for prediction. In our case, in its most general setting, the Price equation is usually formulated in a way that partitions a \emph{given} amount of phenotypic change between two populations (usually but not necessarily the same population at two different times) into change due to selection, transmission bias, etc., rather than \emph{predicting} a trajectory for how much phenotypic change will occur at various future times based on the \emph{current} phenotypic distribution.}~\citep{van_veelen_use_2005,frank_natural_2012,simon_continuous-time_2014,queller_fundamental_2017}. However, this need not be the case if we are willing to compromise slightly - Several authors have put forth more predictive versions of the Price equation by moving to a continuous time differential equation framework in which the Price equation is dynamically sufficient but manifests in a slightly less general form~\citep{page_unifying_2002,lion_theoretical_2018,day_price_2020}.

### 2023-03-23 ind_files/02_math_background.tex
Format citations in epigraphs

\epigraph{\justifying Like most mathematicians, he takes the hopeful biologist to the edge of a pond, points out that a good swim will help his work, and then pushes him in and leaves him to drown.}{Charles Elton, speaking about Lotka~\citep{elton_eppur_1935}}

>>>>

\epigraph{\justifying Like most mathematicians, he takes the hopeful biologist to the edge of a pond, points out that a good swim will help his work, and then pushes him in and leaves him to drown.}{Charles~\citet{elton_eppur_1935}, speaking about Lotka}

### 2023-03-23 ind_files/03_BD_processes.tex
Format citations in epigraphs

\epigraph{\justifying Somewhere [...] between the specific that has no meaning and the general that has no content there must be, for each purpose and at each level of abstraction, an optimum degree of generality}{Kenneth Boulding~\citep{boulding_general_1956}}

>>>>

\epigraph{\justifying Somewhere [...] between the specific that has no meaning and the general that has no content there must be, for each purpose and at each level of abstraction, an optimum degree of generality}{Kenneth~\citet{boulding_general_1956}}

### 2023-03-23 ind_files/04_field_eqns.tex
Format citations in epigraphs

\epigraph{\justifying The result has been forty years of bewilderment about what he meant, whereas if he had been willing to make a slight sacrifice of strict mathematical propriety (as I have done) he could have expressed himself in a way that everyone would have understood}{George Price, speaking about Fisher~\citep{price_fishers_1972}}

>>>>

\epigraph{\justifying The result has been forty years of bewilderment about what he meant, whereas if he had been willing to make a slight sacrifice of strict mathematical propriety (as I have done) he could have expressed himself in a way that everyone would have understood}{George~\citet{price_fishers_1972}, speaking about Fisher}

### 2023-03-23 ind_files/05_unification_discussion.tex
Format citations in epigraphs

\epigraph{\justifying Not only is algebraic reasoning exact; it imposes an exactness on the verbal postulates made before algebra can start which is usually lacking in the first verbal formulations of scientific principles.}{J.B.S. Haldane~\citep{haldane_defense_1964}}

>>>>

\epigraph{\justifying Not only is algebraic reasoning exact; it imposes an exactness on the verbal postulates made before algebra can start which is usually lacking in the first verbal formulations of scientific principles.}{J.B.S.~\citet{haldane_defense_1964}}

### 2023-03-23 ind_files/06_nontechnical_discussion.tex
Format citations in epigraphs

\epigraph{\justifying Writing in the last months of this millenium, it is clear that the prime intellectual task of the future lies in constructing an appropriate theoretical framework for biology}{Sydney Brenner~\citep{brenner_theoretical_1999}}

@@

Descriptions such as the classic Price equation and the equations I present in this thesis `abstract away' system-specific details and almost inevitably come at the cost of precision~\citep{levins_strategy_1966, potochnik_idealization_2018}. These approaches are thus intended to complement empirical studies and modelling approaches that carefully study specific systems and generate vital knowledge about how these systems behave. To quote Robert Millikan~\citep{millikan_electron_1924}, ``Science walks forward on two feet, namely theory and experiment [...] Sometimes it is one foot which is put forward first, sometimes the other, but continuous progress is only made by the use of both - by theorizing and then testing, or by finding new relations in the process of experimenting and then bringing the theoretical foot up and pushing it on beyond, and so on in unending alterations.''

>>>>

\epigraph{\justifying Writing in the last months of this millenium, it is clear that the prime intellectual task of the future lies in constructing an appropriate theoretical framework for biology}{Sydney~\citet{brenner_theoretical_1999}}

@@

Descriptions such as the classic Price equation and the equations I present in this thesis `abstract away' system-specific details and almost inevitably come at the cost of precision~\citep{levins_strategy_1966, potochnik_idealization_2018}. These approaches are thus intended to complement empirical studies and modelling approaches that carefully study specific systems and generate vital knowledge about how these systems behave. To quote Robert~\citet{millikan_electron_1924}, ``Science walks forward on two feet, namely theory and experiment [...] Sometimes it is one foot which is put forward first, sometimes the other, but continuous progress is only made by the use of both - by theorizing and then testing, or by finding new relations in the process of experimenting and then bringing the theoretical foot up and pushing it on beyond, and so on in unending alterations.''
